---
title: "Simulations Pre-Class Project"
date: "Due March 13, 2017 at 5:00pm"
output:
  html_document


---

<style type="text/css">
.table {

    width: 80%;
    margin-left:10%; 
    margin-right:10%;
}
</style>
```{r,setup, echo=FALSE, cache=TRUE}
## numbers >= 10^5 will be denoted in scientific notation,
## and rounded to 2 digits
options(scipen = 3, digits = 3)
```




#Project Goals:


With this project we will simulate a famoues probability problem. This will not require knowledge of probability or statistics but only the logic to follow the steps in order to simulate this problem. This is one way to solve problems by using the computer. 

 1. **Gambler's Ruin**: Suppose you have a bankroll of $1000 and make bets of $100 on a fair game. By simulating the outcome directly for at most 5000 iterations of the game (or hands), estimate:
    a. the probability that you have "busted" (lost all your money) by the time you have placed your one hundredth bet. 
```{r}

#You lose when the cumulative sum (where a win is 1 and a loss is -1) is such that -bankroll/bet >= -cumsum (i.e. bankroll/bet tells you how many net losses you can sustain)
play_n_rounds1 <- function(n, bankroll, bet, p=.5){ #here we simulate n rounds
cumsum <- 0
thisround <- 0
roundsmatrix <- matrix(NA, 1, n, byrow=F) #initiallizing the values and matrix we will be using
  for(i in 1:n) { #want n many rounds
    if(cumsum == -bankroll/bet) { #if your out of money, keep the ratio of your cumsum
      roundsmatrix[i] <- cumsum
    } else { #play again
      thisround <- rbinom(1,1,p)*2-1 #gives positive 1 or negative 1 (a win or a loss)
    cumsum <- cumsum + thisround #cumsum keeps a cumulative sum
    roundsmatrix[i] <- cumsum}
  }
print(roundsmatrix[n]) #if this value is -bankroll/bet you ran out of money, otherwise you made it
}
play_n_rounds1(12, 1000,1000) #testing the function


simulate_survival <- function (sims, n, bankroll, bet, p=.5) {
  results <- matrix(NA, 1,sims, byrow=F) #going to make a matrix store the simulationo results in
  for(i in 1:sims) {
    results[1,i] <- play_n_rounds1(n, bankroll, bet, p)
  } #this gives us our needed output, but it is messy, if you ran out of money the value is -bankroll/bets, otherwise it is a higher value. Let's clean that up now
  for(i in 1:sims) {
    if(results[1,i] == -bankroll/bet) {
      results[1,i] <- 0 #let zero represent people who went bust
    } else (results[1,i] <- 1) #let 1 represent people who made it through without going bust
  }
  return(results)
} 
simulate_survival(100,12, 1000,1000)
 
p_win <- function(matrix) {print(mean(matrix))}
p_lose<- function(matrix) {print(1-mean(matrix))}


#we want to run a simulation with 5000 sims, n of 100, a bankroll of 1000, and a bet of 100, (and a p of 0.5)
p_lose(simulate_survival(5000, 100, 1000, 100))


```
    b. the probability that you have busted by the time you have placed your five hundredth bet by simulating the outcome directly. 
```{r}
p_lose(simulate_survival(5000, 500, 1000, 100)) #this is the same thing, we just changed n
    
```
    c. the mean time you go bust, given that you go bust within the first 5000 hands.
```{r}
#now we need a counter, and we only want to look at the times where we bust, first we will ad a counter into the earlier code then we will  limit ourselves to busts

play_n_rounds2 <- function(n, bankroll, bet, p=.5){ #play n rounds 2 will tell us how many rounds we survived ( with a maximum value of n)
cumsum <- 0
thisround <- 0
count <- 0
roundsmatrix <- matrix(NA, 2, n, byrow=F) #This time we will work with a 2 row matrix (one for the count)
  for(i in 1:n) { 
    if(cumsum == -bankroll/bet) { 
      roundsmatrix[1,i] <- cumsum
      roundsmatrix[2,i] <- count #if you've lost, the count stays the same, lets fill that in for every column thereafter
    } else { 
      count = i #set count as i (note we can't simply use i or we'd have an issue in the above if statement since it'd keep growing even if we ran out of bankroll)
      thisround <- rbinom(1,1,p)*2-1 
    cumsum <- cumsum + thisround 
    roundsmatrix[1,i] <- cumsum
    roundsmatrix[2,i] <- count}
  }
print(roundsmatrix[2,n]) #tells you how many rounds were survived
}
play_n_rounds2(12, 1000,500) #testing the function



simulate_bust_time <- function (sims, n, bankroll, bet, p=.5) { #this time it will store how many rounds were survived in the simulation, not this is very similar to simulate_survival
  results <- matrix(NA, 1,sims, byrow=F) 
  for(i in 1:sims) {
    results[1,i] <- play_n_rounds2(n, bankroll, bet, p) #calls rounds2 not rounds1
  } #now we need to remove anyone who didn't go bust
  for(i in 1:sims) {
    if(results[1,i] == n) {
      results[1,i] <- NA #set people who made it to the end as NA
    } #no else statement needed
  }
  return(results)
} 

mean(simulate_bust_time(500, 5000, 1000, 100), na.rm=T) #ran a simulation of 500 times looking at the average amount of time busters spent playing before they busted (looking only at those who busted within their first 5000 plays) - they started with 1000 and bet 100 per play. Mean was 509.8045
```
    d. the mean and variance of your bankroll after 100 hands (including busts).
```{r}
#we need to convert back from the cumsum to the money value by multiplying it by the bet. This will require changing some of our earlier code which had given a 0 1 output
whats_left <- function(bankroll, bet, cumsum) {
  finalbankroll = bankroll + bet*cumsum
  return(finalbankroll)
}



play_n_rounds3 <- function(n, bankroll, bet, p=.5){ #note this is 3rd one of these
cumsum <- 0
thisround <- 0
roundsmatrix <- matrix(NA, 1, n, byrow=F) 
  for(i in 1:n) { 
    if(cumsum == -bankroll/bet) { 
      roundsmatrix[i] <- cumsum
    } else { 
      thisround <- rbinom(1,1,p)*2-1 
    cumsum <- cumsum + thisround 
    roundsmatrix[i] <- bankroll + bet*cumsum} #We change what we store now so that roundsmatrix (which will be our output) has our dollar total
  }
print(roundsmatrix[n]) #this will be zero if we go bankrupt
}
play_n_rounds3(12, 1000,100) #testing the function


simulate_gambles <- function (sims, n, bankroll, bet, p=.5) { #again it is roughly the same function with some changes
  results <- matrix(NA, 1,sims, byrow=F) 
  for(i in 1:sims) {
    results[1,i] <- play_n_rounds3(n, bankroll, bet, p) #this time we use rounds3
  } 
#no longer need our for loop since we are using all cases
  return(results)
} 
test <- simulate_gambles(1000,100, 1000,100) #1000 simulations of 100 hands with 1000 strting bankroll and bets of 100
mean(test) #gave 1012.99
(sd(test))^2 #not sure why var is throwing an error??? sd works fine though so I'll use it, gave 809090.5


```
    e. the mean and variance of your bankroll after 500 hands (including busts).
```{r}
test2 <- simulate_gambles(1000,500, 1000,100) #1000 simulations of 500 hands with 1000 strting bankroll and bets of 100
mean(test) #gave 1016.49
(sd(test))^2 #gave 2760734
    
```
 
Note: you *must* stop playing if your player has gone bust. How will you handle this in the `for` loop?

2. Repeat the previous problem with betting on black in American roulette, where the probability of winning on any spin is 18/38 for an even payout.

3. **Markov Chains**. Suppose you have a game where the probability of winning on your first hand is 48%; each time you win, that probability goes up by one percentage point for the next game (to a maximum of 100%, where it must stay), and each time you lose, it goes back down to 48%. Assume you cannot go bust and that the size of your wager is a constant $100.
    a. Is this a fair game? Simulate one hundred thousand sequential hands to determine the size of your return. Then repeat this simulation 99 more times to get a range of values to calculate the expectation.
    b. Repeat this process but change the starting probability to a new value within 2% either way. Get the expected return after 100 repetitions. Keep exploring until you have a return value that is as fair as you can make it. Can you do this automatically?
    c. Repeat again, keeping the initial probability at 48%, but this time change the probability increment to a value different from 1%. Get the expected return after 100 repetitions. Keep changing this value until you have a return value that is as fair as you can make it. 


4. Creating a Bootstrap function. There is a particular concept called [bootstrapping]
(https://en.wikipedia.org/wiki/Bootstrapping_(statistics)) where we can easily create 95% confidence intervals, even for complex estimators.

The steps of this process are:

  a. Draw a sample, with replacement, from your data which is the same length of your data.
  b. Calculate the statistic of interest on this boostrap sample (ie mean, variance, regression,...)
  c. Peform steps 1:2 at least 1000 times over until you have a vector of your statistics. 
  d. The lower bound of a 95% CI will be the 0.025 percentile
  e. The upper bound of a 95% CI will be the 0.975 percentile

Make a function called `boot_ci` which calculates the 95% confidence interval in this manner. 

5. For problems 3b and 3c, you calculated a mean value. Because you saved these final results in a vector, use the bootstrap to estimate the variance of the return in each case for your final answer. Once you have these results, which game has the smaller variance in returns?