---
title: "Simulations Pre-Class Project"
date: "Due March 13, 2017 at 5:00pm"
output:
  html_document


---

<style type="text/css">
.table {

    width: 80%;
    margin-left:10%; 
    margin-right:10%;
}
</style>
```{r,setup, echo=FALSE, cache=TRUE}
## numbers >= 10^5 will be denoted in scientific notation,
## and rounded to 2 digits
options(scipen = 3, digits = 3)
```




#Project Goals:


With this project we will simulate a famoues probability problem. This will not require knowledge of probability or statistics but only the logic to follow the steps in order to simulate this problem. This is one way to solve problems by using the computer. 

 1. **Gambler's Ruin**: Suppose you have a bankroll of $1000 and make bets of $100 on a fair game. By simulating the outcome directly for at most 5000 iterations of the game (or hands), estimate:
    a. the probability that you have "busted" (lost all your money) by the time you have placed your one hundredth bet. 
```{r}

#You lose when the cumulative sum (where a win is 1 and a loss is -1) is such that -bankroll/bet >= -cumsum (i.e. bankroll/bet tells you how many net losses you can sustain)
set.seed(1)

play_n_rounds1 <- function(n, bankroll, bet, p=.5){ #here we simulate n rounds
cumsum <- 0
thisround <- 0
bank <- bankroll
roundsmatrix <- matrix(NA, 4, n, byrow=F) #initiallizing the values and matrix we will be using, row 1 contains the amount lost or gained each hand, row 2 will hold number of hands played, row 3 contains the total bankroll amount
  for(i in 1:n) { #want n many rounds, row 4 says whether we went bust or not
    if(cumsum == -bankroll) { #if your out of money, keep the ratio of your cumsum
      roundsmatrix[1, i] <- cumsum
      roundsmatrix[2, i] <- count
      roundsmatrix[3, i] <- bank #should be 0
      roundsmatrix[4, i] <- bank #should be 0
    } else { #play again
      count <- i
      win_loss <- rbinom(1,1,p)*2-1
      thisround <- win_loss*bet #gives us a positive bet or a negative bet
      
      cumsum <- cumsum + thisround #cumsum keeps a cumulative sum
      roundsmatrix[1, i] <- cumsum
    
      roundsmatrix[2, i] <- count
    
      bank <- bank + bet*win_loss
      roundsmatrix[3, i] <- bank}
    
      if (bank == 0) {
        roundsmatrix[4, i] <- 0 #0 indicates that we went bankrupt on the last hand
      } else {
        roundsmatrix[4, i] <- 1 #1 indicates that we have not yet gone bankrupt
      }
      
  }
print(roundsmatrix[,n]) #prints out the entire matrix
}

play_n_rounds1(12, 1000,1000) #testing the function


simulate_survival <- function (sims, n, bankroll, bet, p=.5) {
  results <- matrix(NA, 4, sims, byrow=F) #going to make a matrix store the simulationo results in
  for(i in 1:sims) {
    results[,i] <- play_n_rounds1(n, bankroll, bet, p) #column n in the output (end of the game) gets stored in the results for iteration i
  }
  return(results)
} 
simulate_survival(100,12, 1000,1000)
 
p_win <- function(matrix_row) {print(mean(matrix_row))}
p_lose<- function(matrix_row) {print(1-mean(matrix_row))}


#we want to run a simulation with 5000 sims, n of 100, a bankroll of 1000, and a bet of 100, (and a p of 0.5)
p_lose(simulate_survival(5000, 100, 1000, 100)[4,])


```
    b. the probability that you have busted by the time you have placed your five hundredth bet by simulating the outcome directly. 
```{r}
p_lose(simulate_survival(5000, 500, 1000, 100)[4,]) #this is the same thing, we just changed n
    
```
    c. the mean time you go bust, given that you go bust within the first 5000 hands.
```{r}
#now we need a counter, and we only want to look at the times where we bust, first we will ad a counter into the earlier code then we will  limit ourselves to busts
play_n_rounds2 <- function(n, bankroll, bet, p=.5){ #play n rounds 2 will tell us how many rounds we survived ( with a maximum value of n)
cumsum <- 0
thisround <- 0
count <- 0
roundsmatrix <- matrix(NA, 2, n, byrow=F) #This time we will work with a 2 row matrix (one for the count)
  for(i in 1:n) { 
    if(cumsum == -bankroll/bet) { 
      roundsmatrix[1,i] <- cumsum
      roundsmatrix[2,i] <- count #if you've lost, the count stays the same, lets fill that in for every column thereafter
    } else { 
      count = i #set count as i (note we can't simply use i or we'd have an issue in the above if statement since it'd keep growing even if we ran out of bankroll)
      thisround <- rbinom(1,1,p)*2-1 
    cumsum <- cumsum + thisround 
    roundsmatrix[1,i] <- cumsum
    roundsmatrix[2,i] <- count}
  }
print(roundsmatrix[2,n]) #tells you how many rounds were survived
}
play_n_rounds2(12, 1000,500) #testing the function



simulate_bust_time <- function (sims, n, bankroll, bet, p=.5) { #this time it will store how many rounds were survived in the simulation, not this is very similar to simulate_survival
  results <- matrix(NA, 1,sims, byrow=F) 
  for(i in 1:sims) {
    results[1,i] <- play_n_rounds2(n, bankroll, bet, p) #calls rounds2 not rounds1
  } #now we need to remove anyone who didn't go bust
  for(i in 1:sims) {
    if(results[1,i] == n) {
      results[1,i] <- NA #set people who made it to the end as NA
    } #no else statement needed
  }
  return(results)
} 

mean(simulate_bust_time(500, 5000, 1000, 100), na.rm=T) #ran a simulation of 500 times looking at the average amount of time busters spent playing before they busted (looking only at those who busted within their first 5000 plays) - they started with 1000 and bet 100 per play. Mean was 509.8045
```
    d. the mean and variance of your bankroll after 100 hands (including busts).
```{r}
part_d <- simulate_survival(5000, 100, 1000, 100) #5000 simulations of 100 hands
mean(part_d[3, ])
var(part_d[3, ])
```
    e. the mean and variance of your bankroll after 500 hands (including busts).
```{r}
part_e <- simulate_survival(5000, 500, 1000, 100) #5000 simulations of 500 hands
mean(part_e[3, ])
var(part_e[3, ])
    
```
 
Note: you *must* stop playing if your player has gone bust. How will you handle this in the `for` loop?

2. Repeat the previous problem with betting on black in American roulette, where the probability of winning on any spin is 18/38 for an even payout.
```{r}
#part a: 0.501
p_lose(simulate_survival(5000, 100, 1000, 100, p = 18/38)[4,])

#part b: 0.9162
p_lose(simulate_survival(5000, 500, 1000, 100, p = 18/38)[4,])

#part c: 198.568
mean(simulate_bust_time(5000, 5000, 1000, 100, p = 18/38), na.rm=T)

#part d: mean is 592.16, variance is 570020.5
part_d2 <- simulate_survival(5000, 100, 1000, 100, p = 18/38) #5000 simulations of 100 hands
mean(part_d2[3, ])
var(part_d2[3, ])

#part e: mean is 166.32, variance is 401041.9
part_e2 <- simulate_survival(5000, 500, 1000, 100, p = 18/38) #5000 simulations of 500 hands
mean(part_e2[3, ])
var(part_e2[3, ])

```

3. **Markov Chains**. Suppose you have a game where the probability of winning on your first hand is 48%; each time you win, that probability goes up by one percentage point for the next game (to a maximum of 100%, where it must stay), and each time you lose, it goes back down to 48%. Assume you cannot go bust and that the size of your wager is a constant $100.
    a. Is this a fair game? Simulate one hundred thousand sequential hands to determine the size of your return. Then repeat this simulation 99 more times to get a range of values to calculate the expectation.
```{r}
set.seed(1)
#build the markov function based off the gamblers_ruin function
markov <- function(bet, p, maxrun = 100000){
  n <- 0 #initialize count of hands
  prob <- p
  net <- 0 #net represents net winnings
  while (1 > 0) {
    probval <- runif(1)
    if (probval <= p) { #if win game, add bet value to net value; and increase probability of winning by 1%
      net <- net + bet
      p <- p + 0.01
    }else { #if lose game, subtract bet value from net value; and return probability of winning back to original input
      net <- net - bet
      p <- prob
    }
    n <- n + 1 #increase hand by 1
    if (n == maxrun) { #simulate 100,000 sequential hands
      break
      }
  }
  return(net) #return bankroll value
}

winnings <- function(bet, p, sims) {
  size <- numeric() #initialize size vector
  for (i in 1:sims) { #simulate 100 times
    next_size <- markov(bet, p) #fill the size vector using the markov function
    size <- c(size, next_size)
  }
  return(size)
}

mean(winnings(100, 0.48, 100)) #calculate the mean size of returns
```
This is not a fair game: my mean return is -199260.
    b. Repeat this process but change the starting probability to a new value within 2% either way. Get the expected return after 100 repetitions. Keep exploring until you have a return value that is as fair as you can make it. Can you do this automatically?
```{r}
set.seed(1)
for (j in 1:20) { #loop to make process automatic, intend to have starting value between 46% and 50%
  size = numeric() #initialize size vector
  for(i in 1:100) { #simulate 100 times
    next_size <- markov(100, 0.46+0.002*j, maxrun = 100000) #fill the size vector using the markov function
    size <- c(size, next_size)
  }
  if (mean(size) > 0) {
    print(j)
  }
}

```    
    c. Repeat again, keeping the initial probability at 48%, but this time change the probability increment to a value different from 1%. Get the expected return after 100 repetitions. Keep changing this value until you have a return value that is as fair as you can make it.
```{r}
#modify the markov function by including an increment input
markov_inc <- function(bet, p, inc, maxrun = 100000){
  n <- 0 #initialize count of hands
  prob <- p
  net <- 0 #net represents net winnings
  while (1 > 0) {
    probval <- runif(1)
    if (probval <= p) { #if win game, add bet value to net value; and increase probability of winning by 1%
      net <- net + bet
      p <- p + inc
    }else { #if lose game, subtract bet value from net value; and return probability of winning back to original input
      net <- net - bet
      p <- prob
    }
    n <- n + 1 #increase hand by 1
    if (n == maxrun) { #simulate 100,000 sequential hands
      break
      }
  }
  return(net) #return bankroll value
}

set.seed(1)
for (j in 1:20) { #loop to make process automatic, intend to have starting value between 46% and 50%
  size = numeric() #initialize size vector
  for(i in 1:100) { #simulate 100 times
    next_size <- markov_inc(100, 0.48, 0.01+j/1000, maxrun = 100000) #fill the size vector using the markov function
    size <- c(size, next_size)
  }
  if (mean(size) > 0) {
    print(j)
  }
}
```


4. Creating a Bootstrap function. There is a particular concept called [bootstrapping]
(https://en.wikipedia.org/wiki/Bootstrapping_(statistics)) where we can easily create 95% confidence intervals, even for complex estimators.

The steps of this process are:

  a. Draw a sample, with replacement, from your data which is the same length of your data.
  b. Calculate the statistic of interest on this boostrap sample (ie mean, variance, regression,...)
  c. Peform steps 1:2 at least 1000 times over until you have a vector of your statistics. 
  d. The lower bound of a 95% CI will be the 0.025 percentile
  e. The upper bound of a 95% CI will be the 0.975 percentile

Make a function called `boot_ci` which calculates the 95% confidence interval in this manner.

```{r}
boot_ci <- function(data, fun, iterations = 1000, lowerci = 0.025, upperci = 0.975) { #mandatory inputs are dataset of interest and sample statistic you want to calculate; iterations, lowerci, and upperci all have default values
  sample_stats <- vector(length = iterations) #initialize sample_stats vector
  for (i in 1:iterations) { #repeat as many times as specified...
    temp_sample <- sample(data, length(data), replace = TRUE) #sample with replacement from data 
    sample_stats[i] <- fun(temp_sample)
  }
  lowerbound <- quantile(sample_stats, probs = lowerci) #find the lowerbound value
  upperbound <- quantile(sample_stats, probs = upperci) #find the upperbound value
  return(c(lowerbound, upperbound)) #return the 95% CI
}

testing <- c(1:100)
boot_ci(testing, mean)
```


5. For problems 3b and 3c, you calculated a mean value. Because you saved these final results in a vector, use the bootstrap to estimate the variance of the return in each case for your final answer. Once you have these results, which game has the smaller variance in returns?